sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-03-13 doc/paper-jss/paper-jss.Rmd
Copy edit

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}).
Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{$\sim$ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} [@finley2007; @finley2015] since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix \ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @openblas and PARDISO: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \proglang{R}-\pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

>>>>

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}).
Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
\pkg{sdmTMB} can use many `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde\ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} [@finley2007; @finley2015] since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix&nbsp;\ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (\pkg{openBLAS}: @openblas and \pkg{PARDISO}: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table  &nbsp;\ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \proglang{R}-\pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->
